Import the necessary libraries. 
```{r}
library(httr2)
library(tidyverse)
library(janitor)
library(jsonlite)
library(purrr)
library(dplyr)
library(lubridate)
library(ggplot2)
```

Load in the dataset. 
```{r}
load("../data/WrangledData.rdata")
dat <- get("dat")
head(dat)
```

Create a dataframe called us_cases that combines all the cases for all US states. 
```{r}
us_cases <- dat |>
  mutate(
    year = year(as.Date(date)), 
    month = month(as.Date(date), label = TRUE)
  ) |>
  group_by(year, month) |>
  summarize(
    total_cases = sum(cases, na.rm = TRUE),
    total_population = sum(population, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    date = as.Date(paste(year, month, "1", sep = "-"), format = "%Y-%b-%d")
  ) |>
  mutate(cases_per_100k = (total_cases / total_population) * 100000) |>
  drop_na(cases_per_100k)
head(us_cases,2)
```

Plot the total COVID-19 cases by month/year in the US. 
```{r}
plot <- ggplot(us_cases, aes(x = date, y = cases_per_100k)) +
  geom_line() +
  labs(
    title = "Total COVID-19 Cases by Month in the US",
    x = "Date",
    y = "Total Cases Per 100K"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

ggsave("../docs/total_covid_us.png", plot = plot, width = 20, height = 12, dpi = 300)
```

Now, we can see that it matches the general trend for COVID-19 cases and the deaths per state and region, with fluctuations. 
```{r}
dat_long <- dat |>
  mutate(
    case_rate = (cases / population) * 100000,
    death_rate = (deaths / population) * 100000
  ) |>
  pivot_longer(
    cols = c(case_rate, death_rate),
    names_to = "rate_type",
    values_to = "rate"
  )

plot2 <- ggplot(dat_long, aes(x = date, y = rate, group = state, color = region_name)) +
  geom_line() +
  facet_wrap(~rate_type, scales = "free_y", ncol = 1) +
  labs(
    title = "Trends for COVID-19 Case and Death Rates",
    x = "Date",
    y = "Rate per 100,000 People",
    color = "Region"
  ) +
  theme_minimal()

ggsave("../docs/covid_state_region.png", plot = plot2, width = 20, height = 12, dpi = 300)
```

Based on the graph above, we have decided to define the following time periods:

1. 2020-01-25 to 2021-02- - initial global outbreak 
2. 2021-03-01 to 2021-11-30 - Delta variant surge
3. 2021-12-01 to 2023-04-29 - Omicron wave causing a sharp spike in cases
